Einstein's gravity from a polynomial affine model

### 2015-10-21 AffGR-PRD.tex
Reviewed until Eq. (14)

\documentclass[aps,prd,12pt,twocolumn,superscriptaddress,showpacs,showkeys,reprint]{revtex4-1}

@@

%% --------- Abstract
\begin{abstract}
  We show that the effective field equations for a recently formulated polynomial affine model of gravity, in the sector of a torsion-free connection, accept general Einstein manifolds ---with or without cosmological constant--- as solutions. Moreover, these effective field equations coincide with those obtained from a gravitational Yang--Mills theory known as Stephenson--Kilmister--Yang theory. Additionally, we find a generalization of a minimally coupled massless scalar field in General Relativity with a ``minimally'' coupled scalar field in this affine model.
\end{abstract}

@@

Classically, General Relativity is the most successful theory of gravitational interactions. However, it is believed to be the effective theory of a, yet to be discovered, more fundamental one. Among the problems of General Relativity we mention a few aspects related with the quantization: First, the fact that in the standard formalism of quantization General Relativity is non-renormalizable~\cite{'tHooft:1973us,'tHooft:1974bx,Deser:1974cz,Deser:1974cy}. Second, the Wheeler--DeWitt equations are not well-defined in general due to their nonpolynomial dependence on the Arnowitt--Deser--Misner (or ADM) variables~\cite{Arnowitt:1959ah,Arnowitt:1960es,WheelerGeo,DeWitt:1967yk,DeWitt:1967ub,DeWitt:1967uc}. One of the most successful quantization methods of gravitational interactions ---developed from ideas by Gambini and Trias~\cite{Gambini:1980yz,Gambini:1986ew}, and Ashtekar~\cite{Ashtekar:1986yd,Ashtekar:1987gu}---, hints that General Relativity is not as complete as expected, since the Lagrangian formulation includes (at least) an extra term known as Holst term~\cite{Holst:1995pc}. On the other hand, it is well known that in the standard formulation of General Relativity the skew-symmetric part of the connection (or torsion) is assumed to be zero, albeit it is known that such assumption can be relaxed once matter is added, in particular fermionic matter~\cite{Kibble:1961ba}.

@@

Therefore, under the premises of incompleteness and non-polynomial structure of the theory, 
%%%%%%%%% THE MODEL %%%%%%%%%
in a previous paper~\cite{Skirzewski:2014eta}, we presented a polynomial (purely) affine gravitational model in four dimensions built up entirely on the basis of full diffeomorphism invariance, and restricting ourselves to consider only naive power-counting renormalizable terms (which by no means guarantees the renormalizability of the system). We showed that its non-relativistic limit around a homogeneous and isotropic spacetime yields a Newtonian gravitational potential, despite the existence of torsion. We also highlighted the possibility of using standard methods to quantize the model, due to its polynomial structure (unlike the earliest proposals~\cite{Eddington1923math,schrodinger1950space}), and also the likelihood of avoiding the uniqueness of diffeomorphism-invariant states from Loop Quantum Gravity programme~\cite{Lewandowski:2005jk}, due to the absence of a fundamental metric, or in other words, the absence of flux operators.

@@

Using the above decomposition, the most general (power-counting renormalizable) action preserving diffeomorphisms is~\footnote{We have dropped all terms which can be related to those in Eq.~\eqref{4dfull} through a boundary term.}

@@

@@

From the geodesic equations and the linearized action on scalar perturbation, we found that only a few terms in the scalar perturbation affect the geodesic equation~\footnote{We cross-check our result using the software Cadabra~\cite{Peeters2007550,peeters2007symbolic,Peeters:2007wn}.}. The effective geodesic deviation is given by
\begin{equation}
  \label{NewtonPot}
  \Gamma^i{}_{00} = - \frac{1}{8\pi} \frac{ \partial\mathcal{L}_{\text{Matter}} }{ \partial \mathfrak{g}^{00} } \, \nabla^i \left(\frac{1}{|\vec{x}|}\right),
\end{equation}
is the usual Newtonian force induced by a massive source.

\section{\label{rlimit}Relativistic limit}
%%%%%%%%% COMPARISON GR %%%%%%%%%
Despite we obtained the correct Newtonian gravitational potential in the non-relativistic limit, see Eq.~\eqref{NewtonPot}, the post-Newtonian approximation is necessary to explain gravitational phenomena like Mercury's perihelion and the deviation of light by a gravitational source. Therefore, we should go beyond the Newtonian limit in order to compare our model with General Relativity.

%%\hl{A comparison between our affine model and GR requires the imposition of vanishing torsion (JZ: Why?).}
Even though a comparison between the models does not require the imposition of vanishing torsion, for the sake of simplicity, at this stage we shall focus on a sector of the theory in which the connection is torsion-free. Notice that the vanishing torsion limit ---equivalent to take $T^{\lambda,\mu\nu} \to 0$ and $A_\mu \to 0$--- cannot be taken at the action level, and should be taken in the field equations. Interestingly, the field equations from Eq.~\eqref{4dfull} can be consistently \emph{truncated} under such requirements. Since only the terms of the action linear in these fields will be relevant, one can consider the effective action linear in the torsion's fields, i.e.,
\begin{equation}
  %\label{eff-action}
  S_{\text{eff}} = \int\dn{4}{x} \Big( C_1\, R_{\lambda\mu}{}^{\lambda}{}_\nu \nabla_\rho %T^{\nu,\rho\sigma}
  + C_2 \, R_{\mu\rho}{}^{\lambda}{}_\lambda \nabla_\nu \Big) T^{\nu,\mu\rho} ,
\end{equation}
and the only nontrivial field equation after the limit will be the one for the Curtright, $T^{\nu,\mu\rho}$,
\begin{equation}
  \nab{[\rho} R_{\mu]\nu} + \kappa \nab{\nu} R_{\mu\rho}{}^\lambda{}_\lambda = 0,
\end{equation}
with $\kappa$ a constant related with the original couplings of the model. We assume that the connection is compatible with volume form, i.e., it is equi-affine~\cite{nomizu1994affine,MO-Bryant02}.  The equi-affine condition assures that the Ricci tensor of the connection is symmetric, and the contraction of the last indices vanishes, thus the second term in the field equation is absent and the gravitational equations are
\begin{equation}
  \nab{[\rho} R_{\mu]\nu} = 0.
  \label{SimpleEOM}
\end{equation}

Equation~\eqref{SimpleEOM} is a generalization of a condition known in Riemannian geometry as covariantly constant Ricci curvature ---aka parallel Ricci curvature---, \mbox{$\nab{\rho} R_{\mu\nu} = 0$.}  All Einstein manifolds, whose Ricci tensor is proportional to the metric, \mbox{$R_{\mu\nu} \propto g_{\mu\nu}$,} satisfy the parallel Ricci condition due to the metricity condition, and therefore every vacuum solution to the Einstein's equations solves the (torsion-free) field equations of our model. Consequently, the fact that the non-relativistic limit of the gravitational potential in Eq.~\eqref{NewtonPot} yields a Newtonian potential seems clearer, and we can argue that even the post-Newtonian corrections coming from General Relativity are present in the chosen scenario of our model. As an additional comment, the parallel Ricci curvature assures that although the manifold is not in general a Einstein manifold, the metric has to be locally a product of Einstein metrics~\cite{Besse}.

>>>>

\documentclass[aps,prd,12pt,twocolumn,superscriptaddress,showpacs,showkeys,reprint,longbibliography]{revtex4-1}

@@

%% --------- Abstract
\begin{abstract}
  We show that the effective field equations for a recently formulated polynomial affine model of gravity, in the sector of a torsion-free connection, accept general Einstein manifolds ---with or without cosmological constant--- as solutions. Moreover, these effective field equations coincide with some of those obtained from a gravitational Yang--Mills theory known as Stephenson--Kilmister--Yang theory. Additionally, we find a generalization of a minimally coupled massless scalar field in General Relativity with a ``minimally'' coupled scalar field in this affine model. Finally, we present a naive analysis of the propagators of the gravitational theory, an their relation with the presence of ghost in the model. 
\end{abstract}

@@

Classically, General Relativity is the most successful theory of gravitational interactions. However, it is believed to be the effective theory of a, yet to be discovered, more fundamental one. Among the problems of General Relativity we mention a few aspects related with the quantization: First, the fact that in the standard formalism of quantization General Relativity is non-renormalizable~\cite{'tHooft:1973us,'tHooft:1974bx,Deser:1974cz,Deser:1974cy}. Second, the Wheeler--DeWitt equations are not well-defined in general due to their nonpolynomial dependence on the Arnowitt--Deser--Misner (or ADM) variables~\cite{Arnowitt:1959ah,Arnowitt:1960es,WheelerGeo,DeWitt:1967yk,DeWitt:1967ub,DeWitt:1967uc}. One of the most successful quantization methods of gravitational interactions ---developed from ideas by Gambini and Trias~\cite{Gambini:1980yz,Gambini:1986ew}, and Ashtekar~\cite{Ashtekar:1986yd,Ashtekar:1987gu}---, hints that General Relativity is not as complete as expected, since the Lagrangian formulation includes (at least) an extra term known as Holst term~\cite{Holst:1995pc}. On the other hand, it is well known that in the standard formulation of General Relativity the skew-symmetric part of the connection (or torsion) is assumed to be zero, albeit it is known that such assumption can be relaxed once matter is added, in particular fermionic matter~\cite{Kibble:1961ba,Hehl:1976kj,Shapiro:2001rz,Hammond:2002rm}.

@@

%%%%%%%%% THE MODEL %%%%%%%%%
In order to undertake the premises of incompleteness and non-polynomial structure of the theory, in a previous paper~\cite{Skirzewski:2014eta}, we presented a polynomial (purely) affine gravitational model in four dimensions built up entirely on the basis of full diffeomorphism invariance. Worth to mention that the most general action built this way contains only power-counting renormalizable terms (which by no means guarantees the renormalizability of the system). We showed that its non-relativistic limit around a homogeneous and isotropic spacetime yields a Newtonian gravitational potential, despite the existence of torsion. We also highlighted the possibility of using standard methods to quantize the model, due to its polynomial structure (unlike the earliest proposals~\cite{Eddington1923math,schrodinger1950space}), and also the likelihood of avoiding the uniqueness of diffeomorphism-invariant states from Loop Quantum Gravity programme~\cite{Lewandowski:2005jk}, due to the absence of a fundamental metric, or in other words, the absence of flux operators.

@@

Using the above decomposition, the most general action preserving diffeomorphisms is

@@

In Eq.~\eqref{4dfull}, we have dropped all terms which can be related through partial integration, and those which are topological invariant. A straightforward dimensional counting shows the impossibility of other contributions in to the action, \hl{see Appendix ...}. Interestingly, the above action turns to be power-counting renormalizable, which does not guarantees renormalizability but is a good signal. The structure of the model yields no three-point graviton vertices, which might allow to overcome  the no-go theorems found in Refs.~\cite{McGady:2013sga,Camanho:2014apa}. It is worth mentioning that all coupling constants are dimensionless, which might be a hint of conformal invariance of the model~\cite{Buchholz:1976hz}.

@@

From the geodesic equations and the linearized action on scalar perturbation, we found that only a few terms in the scalar perturbation affect the geodesic equation~\footnote{We cross-check our result using the software Cadabra~\cite{Peeters2007550,peeters2007symbolic,Peeters:2007wn}.}. The effective geodesic deviation is given by
\begin{equation}
  \label{NewtonPot}
  \Gamma^i{}_{00} = - \frac{1}{8\pi} \frac{ \partial\mathcal{L}_{\text{Matter}} }{ \partial \mathfrak{g}^{00} } \, \nabla^i \left(\frac{1}{|\vec{x}|}\right),
\end{equation}
is the usual Newtonian force induced by a massive source. Notice that we restricted ourselves to consider $\mathfrak{g}^{\mu \nu} = \bar{g}^{\mu\nu}$ to obtain the above result.

\section{\label{rlimit}Relativistic limit}
%%%%%%%%% COMPARISON GR %%%%%%%%%
Despite we obtained the correct Newtonian gravitational potential in the non-relativistic limit, see Eq.~\eqref{NewtonPot}, a beyond-Newtonian approximation is necessary to explain gravitational phenomena like Mercury's perihelion and the deviation of light by a gravitational source. Therefore, we should go beyond the Newtonian limit in order to compare our model with General Relativity.

%%\hl{A comparison between our affine model and GR requires the imposition of vanishing torsion (JZ: Why?).}
Even though a comparison between the models does not require the imposition of vanishing torsion, for the sake of simplicity, at this stage we shall focus on a sector of the theory in which the connection is torsion-free. Notice that the vanishing torsion limit ---equivalent to take $T^{\lambda,\mu\nu} \to 0$ and $A_\mu \to 0$--- cannot be taken at the action level, and should be taken in the field equations. Interestingly, the field equations from Eq.~\eqref{4dfull} can be consistently \emph{truncated} under such requirements. It can be realized as follows, since only the terms of the action linear in these fields will be relevant, one can consider the effective action linear in the torsion's fields, i.e.,
\begin{equation}
  %\label{eff-action}
  S_{\text{eff}} = \int\dn{4}{x} \Big( C_1\, R_{\lambda\mu}{}^{\lambda}{}_\nu \nabla_\rho %T^{\nu,\rho\sigma}
  + C_2 \, R_{\mu\rho}{}^{\lambda}{}_\lambda \nabla_\nu \Big) T^{\nu,\mu\rho} ,
\end{equation}
and the only nontrivial field equation after the limit will be the one for the Curtright, $T^{\nu,\mu\rho}$,
\begin{equation}
  \nab{[\rho} R_{\mu]\nu} + \kappa \nab{\nu} R_{\mu\rho}{}^\lambda{}_\lambda = 0,
\end{equation}
with $\kappa$ a constant related with the original couplings of the model. We assume that the connection is compatible with volume form, i.e., it is equi-affine~\cite{nomizu1994affine,MO-Bryant02}.  The equi-affine condition assures that the Ricci tensor of the connection is symmetric, and the contraction of the last indices vanishes, thus the second term in the field equation is absent and the gravitational equations are
\begin{equation}
  \nab{[\rho} R_{\mu]\nu} = 0.
  \label{SimpleEOM}
\end{equation}

Equation~\eqref{SimpleEOM} is a generalization of a condition known in Riemannian geometry as covariantly constant Ricci curvature ---aka parallel Ricci curvature---, \mbox{$\nab{\rho} R_{\mu\nu} = 0$.}  All Einstein manifolds, whose Ricci tensor is proportional to the metric, \mbox{$R_{\mu\nu} \propto g_{\mu\nu}$,} satisfy the parallel Ricci condition due to the metricity condition, and therefore every vacuum solution to the Einstein's equations solves the (torsion-free) field equations of our model. Consequently, the fact that the non-relativistic limit of the gravitational potential in Eq.~\eqref{NewtonPot} yields a Newtonian potential seems clearer, and we can argue that even the post-Newtonian corrections coming from General Relativity are present in the chosen scenario of our model. An additional comment, the parallel Ricci curvature assures that although the manifold is not Einstein in general, the metric has to be locally a product of Einstein metrics~\cite{Besse}.
